Future challenges for Research Software Engineering

### 2025-05-15 future_challenges_for_RSEng.tex
Remaining updates from proof read of paper.

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{}{career-paths}
We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units.\cite{Kempf2025-draft}
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organisation.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory, but in keeping the quantum computers humming. He became the go-to expert for troubleshooting entangled qubits and stabilizing delicate cryogenic systems. His unique blend of physics knowledge and hands-on problem solving earned him a new title among his peers: the first Quantum RTP. In a world where theory meets reality, Ken quietly ensures the future of quantum computing stays on track.
\end{story}
\section{What is in this future for RSEs and where to go from here?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
It is hard to conclude a piece written about the future, but we have tried to list some of the trends that we foresee.
The material presented here, is based on a workshop, bt extended with the author's ideas.
Nevertheless there are some things that we declared out of scope here.
Among that is the digital transformation in the wider society.
Digital competencies in the general population continue to increase, making jumping to software development easier than before.
But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.
We have also not delved into the interaction with industry much.
There are also some trends that believe to stay with us.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, that arise from technologies such as generally available AI systems.

We hope that the standard of research software has risen across the board to enable an easy reuse in new research.
We hope that a cultural change is taking place to recognize research software as the integral part of science that it is
within broad parts of academia and is supported by political actors.
To conclude, RSEs will very likely be doing different work
than what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
beings -- will stay the same.
They will still develop software artifacts together with scientists,
since better software leads to better research,
better research leads to more better software.

>>>>

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{}{career-paths}
We would like to see the emergence of more dedicated personel, i.e. other (digital) RTPs next to the RSE.
Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units.\cite{Kempf2025-draft}
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding can more difficult to acquire in lower tiers of an of a research organisation (i.e. at departmental or individual research group level).
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg and Göttingen, as well as several universities in the UK where the concept of dedicated RSE groups began to emerge around 2013).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups prove themselves to be increasingly valuable and, hence, sustainable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory, but in keeping the quantum computers humming. He became the go-to expert for troubleshooting entangled qubits and stabilizing delicate cryogenic systems. His unique blend of physics knowledge and hands-on problem solving earned him a new title among his peers: the first Quantum RTP. In a world where theory meets reality, Ken quietly ensures the future of quantum computing stays on track.
\end{story}
\section{What is in this future for RSEs and where to go from here?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
It is hard to conclude a piece written about the future, but we have tried to list some of the trends that we foresee.
The material presented here is based on the outputs of a workshop, but extended with the authors' ideas.
Nevertheless there are some things that we declared out of scope here.
Among that is the ongoing digital transformation in wider society.
Digital competencies in the general population continue to increase, making jumping to software development easier than before.
But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.
We have also not delved into the interaction with industry much.
There are also some trends that we believe will stay with us over the coming years.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, that arise from technologies such as generally available AI systems.

We hope that the standard of research software will have risen across the board, enabling easier reuse of prior software outputs in new research.
We also hope that a cultural change is taking place to recognize research software as an integral part of science that it is
widely recognized and rewarded across academia and is supported by political actors.
To conclude, RSEs will very likely be doing different work
to what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
beings -- will stay the same.
They will still develop software artifacts together with scientists,
since better software leads to better research,
and better research leads to better software.
